---
title: "Practical Machine Learning Course Project Report"
author: "Eduardo"
date: "1st November 2016"
output: html_document
---


Data Processing
---------------
Set libraries used in this analysis
```{r loadLibraries, warning=FALSE, results="hide", message=FALSE}
library(caret)
library(rpart.plot)
library(rattle)
library(AppliedPredictiveModeling)
library(randomForest)
```

First, the training data is downloaded from the following URL:
```{r}
#url_raw_training <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
file_dest_training <- "pml-training.csv"
#download.file(url=url_raw_training, destfile=file_dest_training, method="curl")
#url_raw_testing <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
file_dest_testing <- "pml-testing.csv"
#download.file(url=url_raw_testing, destfile=file_dest_testing, method="curl")
```

Import the data treating empty values as NA.
```{r}
df_training <- read.csv(file_dest_training, na.strings=c("NA",""), header=TRUE)
colnames_train <- colnames(df_training)
df_testing <- read.csv(file_dest_testing, na.strings=c("NA",""), header=TRUE)
colnames_test <- colnames(df_testing)
```

Verify that the column names (excluding classe and problem_id) are identical in the training and test set.
```{r}
all.equal(colnames_train[1:length(colnames_train)-1], colnames_test[1:length(colnames_train)-1])
```

Cleaning data
-------------
Count the number of non-NAs in each col:
```{r}
nonNAs <- function(x) {
    as.vector(apply(x, 2, function(x) length(which(!is.na(x)))))
}
```

Build vector of missing data or NA columns to drop.
```{r}
colcnts <- nonNAs(df_training)
drops <- c()
for (cnt in 1:length(colcnts)) {
    if (colcnts[cnt] < nrow(df_training)) {
        drops <- c(drops, colnames_train[cnt])
    }
}
```

Drop NA data and the first 7 columns as they're unnecessary for predicting.
```{r}
df_training <- df_training[,!(names(df_training) %in% drops)]
df_training <- df_training[,8:length(colnames(df_training))]

df_testing <- df_testing[,!(names(df_testing) %in% drops)]
df_testing <- df_testing[,8:length(colnames(df_testing))]
```

Show remaining columns.
```{r}
colnames(df_training)

colnames(df_testing)
```

First, check for covariates that have virtually no variablility.
```{r}
nsv <- nearZeroVar(df_training, saveMetrics=TRUE)
nsv
```
Given the results obtained, there's no need to eliminate any covariates due to lack of variablility.

Algortihm
---------

Partioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:
```{r}
inTrain <- createDataPartition(y=df_training$classe, p=0.6, list=FALSE)
myTraining <- df_training[inTrain, ]; myTesting <- df_training[-inTrain, ]
dim(myTraining); dim(myTesting)
```

Using ML algorithms for prediction: Decision Tree
-------------------------------------------------

```{r}

modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
```

Using confusion Matrix to test results:
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```


Using ML algorithms for prediction: Random Forests
--------------------------------------------------
```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```
Predicting in-sample error:
```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
```
Using confusion Matrix to test results:
```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```
The results from the Random Forests are much better than the results from the decision tree.

Generating Files to submit as answers for the Assignment:
--------------------------------------------------------
Finally, using the provided Test Set out-of-sample error.
For Random Forests we use the following formula, which yielded a much better prediction in in-sample:
```{r}
predictionsB2 <- predict(modFitB1, df_testing, type = "class")
```
Function to generate files with predictions to submit for assignment
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictionsB2)
```
